---
title: "Analysis of a TCGA RNA-seq data set on Lung Squamous Cell Carcinoma"
author:
- name: Juan Luis Melero Ollonarte
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: jmelero611@gmail.com
- name: Marta Sanvicente García
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: ?????
date: "`r format(Sys.time(), '%B %e, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

# Introduction

Lung cancer accounts for more deaths than any other cancer in both men and women, about 28\% of all cancer deaths. The prognosis for lung cancer is poor. The Cancer Genome Atlas (TCGA) has samples of the most common type of lung cancer called non-small cell lung cancers. Specifically, the subtypes being studied are called lung adenocarcinoma and lung squamous cell carcinoma. Here we analyze the expression profiles of those
patients, accessible in the form of a raw RNA-seq counts using a pipeline based on the R/Bioconductor software package `r Biocpkg("Rsubread")`.





```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysis.knit.md"), sep="\n")
```












