---
output:
  BiocStyle::html_document
---


```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)

opts_chunk$set(cache=TRUE,
               cache.path="cache/QA",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")
```

# Data import

We start importing the table of counts from the SummarizedExperiment container that is already paired for the previous filtering process. The structure is the same: the rows represent genes and the columns represent samples. 

```{r, message=FALSE}
library(SummarizedExperiment)

paired_se <- readRDS(file.path("results", "paired_seLUSC.rds"))
paired_se
```

The row table of counts contains RNA-seq data from 20115 genes and 94 samples, which come from 47 patients.

We explore the column, that corresponds to samples. It contains the phenotypic data, which in this case corresponds to clinical variables, and their corresponding metadata. 


```{r}
dim(colData(paired_se))
colData(paired_se)[1:5, 1:5]
mcols(colData(paired_se), use.names=TRUE)
```

These metadata consists of two columns of information about the clinical variables. One called `labelDescription` contains a succint description of the variable, often not more self-explanatory than the variable name itself, and the other called 'CDEID' corresponds to the `Common Data Element (CDE)` identifier. This identifier can be useed to search for further information about the associated clinical variable.

Now, we explore the row (feature) data.

```{r}
rowData(paired_se)
rowRanges(paired_se)
```

We are going to compare normal samples with tumor samples. Exploring the metadata, we can see that we have 47 normal samples and 47 tumor samples. As they are paired, the number must be the same.

```{r}
table(paired_se$type)
```

# Quality assessment

Normalitzation consists in the adjustment for sample and gene specific factors, to make gene expression values comparable across samples. This process is really important due to the fact that the samples may have been sequenced at different depth and that there may be sample specific biases related to technical differences in samples extarction, preparation and sequencing. 

The normalitzation is done in two steps: 
* Within-sample: adjustments to compare across features in a sample.
  *Scaling: using counts per million reads (CPM) mapped to the genome.
*Between-sample: adjustments to compare a feature across samples.
  *Sample-specific normalization factors: using the TMM algorithm from the R/Bioconductor package edgeR.
  *Quantile normalization: using the CQN algorithm from the R/Bioconductor package cqn or in the limma-voom pipeline.

To perform quality assessment and normalization we need first to load the [edgeR](http://bioconductor.org/packages/edgeR) R/Bioconductor package and create a `DGEList' object.

```{r, message=FALSE}
library(edgeR)

paired_dge <- DGEList(counts=assays(paired_se)$counts, genes=mcols(paired_se))
names(paired_dge)
saveRDS(paired_dge, file.path("results", "paired_dge.rds"))
```

Once the 'DGEList' object is created, we can performe the scaling to CPM values. Therefore, $\log_2$ CPM values of expression are calculated and used as an additional assay element to ease their manipulation. $\log_2$ CPM units separate better high and low expression, than raw counts or non-logged CPM units.

```{r}
assays(paired_se)$logCPM <- cpm(paired_dge, log=TRUE, prior.count=0.5)
assays(paired_se)$logCPM[1:5, 1:5]
```

## Sequencing depth

Let's examine the sequencing depth in terms of total number of sequence read counts
mapped to the genome per sample. Figure \@ref(fig:libsizes) below shows the
sequencing depth per sample, also known as library sizes, in increasing order.

<!---
you can control the height and width in pixels of the figure with 'out.height' and
'out.width'. Figures are automatically numbered, to refer to them in the main test
you should use the notation shown above as \@ref(fig:xxxx) with xxxx being the label
in the code chunk that also gives the filename of the figure. This name must be unique
--->

```{r libsizes, echo=FALSE, out.width="800px", out.height="600px", fig.cap="Library sizes in increasing order."}
ord <- order(paired_dge$sample$lib.size/1e6)
barplot(paired_dge$sample$lib.size[ord], las=1, ylab="Millions of reads",
                xlab="Samples", col=c("blue", "red")[(paired_se$type[ord] == "tumor") + 1])
legend("topleft", c("normal", "tumor"), fill=c("red", "blue"), inset=0.01)
```

There is the same number of tumor samples and normal samples and they seem to be randonly distributed. However, there is still high variability in the library size.

## Distribution of expression levels among samples

Let's look at the distribution of expression values per sample in terms of logarithmic CPM units. Due to the large number of samples, we display tumor and normal samples separately, and are shown in Figure \@ref(fig:distRawExp)

```{r distRawExp, echo=FALSE, fig.height=4, fig.width=10, out.width="800px", fig.cap="Non-parametric density distribution of expression profiles per sample.", message=FALSE}
library(geneplotter)
par(mfrow=c(1, 2))
multidensity(as.list(as.data.frame(assays(paired_se[, paired_se$type == "tumor"])$logCPM)),
                          xlab="log 2 CPM", legend=NULL, main="Tumor samples", las=1)
multidensity(as.list(as.data.frame(assays(paired_se[, paired_se$type == "normal"])$logCPM)),
                          xlab="log 2 CPM", legend=NULL, main="Normal samples", las=1)
```

We do not appreciate substantial differences between the samples in the distribution of expression values.


## Distribution of expression levels among genes

Let's calculate now the average expression per gene through all the samples. Figure \@ref(fig:exprdist) shows the distribution of those values across genes.

```{r exprdist, echo=FALSE, out.width="400px", fig.cap="Distribution of average expression level per gene."}
avgexp <- rowMeans(assays(paired_se)$logCPM)
hist(avgexp, xlab="log2 CPM", main="", las=1, col="steelblue")
abline(v=1, col="red", lwd=2)
```

RNA-seq expression profiles from lowly-expressed genes can lead to artifacts in downstream differential expression analyses. For this reason, it is common practice to remove them following some criteria, such as: filter out genes below a minimum average CPM (or log2 CPM) value throughout the samples or filter out genes with fewer than a given number of samples meeting a minimum CPM (or log2 CPM) cutoff. This graphic shows what would be the minimum average CPM (red line).


We can also make an MA-plot to see biases due to expression. First, we define the groups that we want to compare. In our case, we use the sample type to define groups, by modifying the DGEList object as follows:

```{r}
paired_dge$samples$group <- paired_se$type
table(paired_dge$samples$group)
```

Here we generate the MA-plot using type sample as grouping factor. 

```{r smear1, echo=FALSE, out.width="400px", fig.cap="MA-plot using type grouping."}
plotSmear(paired_dge, lowess = TRUE, las = 1, cex.lab = 1.5, cex.axis = 1.2)
abline(h = 0, col = "blue", lwd = 2)
```

Figure \@ref(fig:smear1) shows us the need of remove the lower experssed genes to normalize the samples.


## Filtering of lowly-expressed genes

In the light of this plot, we may consider a cutoff of 1 log CPM unit as minimum value of expression to select genes being expressed across samples. Using this cutoff we proceed to filter out lowly-expressed genes.

```{r}
mask <- avgexp > 1
dim(paired_se)
paired_se.filt <- paired_se[mask, ]
dim(paired_se.filt)
paired_dge.filt <- paired_dge[mask, ]
dim(paired_dge.filt)
```

After this filtering process, we end up with 11866 genes.

Store un-normalized versions of the filtered expression data.

```{r}
saveRDS(paired_se.filt, file.path("results", "paired_se.filt.unnorm.rds"))
saveRDS(paired_dge.filt, file.path("results", "paired_dge.filt.unnorm.rds"))
```

We can also use a second approach to filter data using the CPM cutoff value of expression. We will keep only genes that have this minimum level of expression in at least as many samples as the smallest group of comparison. We are still comparing sample type. 

```{r}
cpmcutoff <- round(10/min(paired_dge$sample$lib.size/1e+06), digits = 1)
cpmcutoff
nsamplescutoff <- min(table(paired_se$type))
nsamplescutoff
```

After knowing these parameters we can proceed to mask the lower-expressed genes. 

```{r}
mask2 <- rowSums(cpm(paired_dge) > cpmcutoff) >= nsamplescutoff
paired_se.filt2 <- paired_se[mask2, ]
paired_dge.filt2 <- paired_dge[mask2, ]
dim(paired_se)
dim(paired_se.filt)
dim(paired_se.filt2)
```

After this second kind of filtering, we end up with 14200. As we can see, the first filter is more stringent thant the second one.

Store un-normalized versions of the filtered expression data.

```{r}
saveRDS(paired_se.filt2, file.path("results", "paired_se.filt2.unnorm.rds"))
saveRDS(paired_dge.filt2, file.path("results", "paired_dge.filt2.unnorm.rds"))
```

We will compare both approaches done before in order to compare them and see which one discards more genes and is more restrictive.

```{r exprdist2, echo=FALSE, out.width="400px", fig.cap="Distribution of average expression level per gene and filtering comparative."}
par(mar = c(4, 5, 1, 1))
h <- hist(avgexp, xlab="log2 CPM", main="", 
    las=1, col ="steelblue", cex.axis=1.2, cex.lab=1.5)
x <- cut(rowMeans(assays(paired_se.filt)$logCPM), breaks=h$breaks)
lines(h$mids-0.1, table(x), type = "h", lwd=5, lend=1, col = "olivedrab")
y <- cut(rowMeans(assays(paired_se.filt2)$logCPM), breaks=h$breaks)
lines(h$mids+0.1, table(y), type = "h", lwd=5, lend=1, col = "lightgoldenrod")
legend("topright", c("All genes", "Filtered genes", "Filtered 2 genes"), fill=c("steelblue", "olivedrab", "lightgoldenrod"))
```

After comparing the different approaches used in the filter of low expressed genes, we have decided to contunue working with the first set of filtered genes because this is more restrictive. We can see in Figure \@ref(fig:exprdist2) that the second approach is more conservative with our dataset and even genes with negative logCPM passes through this filter. 


## Normalization

We calculate now the normalization factors on the filtered expression data set.

```{r}
paired_dge.filt <- calcNormFactors(paired_dge.filt)
```

Replace the raw log2 CPM units in the corresponding assay element of the `SummarizedExperiment`
object, by the normalized ones.

```{r}
assays(paired_se.filt)$logCPM <- cpm(paired_dge.filt, log=TRUE, normalized.lib.sizes=TRUE, prior.count=0.25)
```

Store normalized versions of the filtered expression data.

```{r}
saveRDS(paired_se.filt, file.path("results", "paired_se.filt.rds"))
saveRDS(paired_dge.filt, file.path("results", "paired_dge.filt.rds"))
```


## MA-plots

We examine now the MA-plots of the normalized expression profiles. We look first to the tumor samples in Figure \@ref(fig:maPlotsTumor).

<!---
Here we make a MA-plot for each sample. The options 'fig.height' and 'fig.width'
control the relative image size in *inches*. The final image size results from
'height'x'dpi' and 'width'x'dpi', where 'dpi' is the image resolution in
"dots per inch" (by default dpi=72). To scale the image to a desired size use
'out.width' and 'out.height'. More information at http://yihui.name/knitr/options
--->

First, we define the groups that we want to compare. In our case, we use the sample type to define groups, by modifying the DGEList object as follows:

```{r}
paired_dge$samples$group <- paired_se$type
table(paired_dge$samples$group)
paired_dge.filt$samples$group <- paired_se$type
```

Here we generate the MA-plot using type sample as grouping factor. 

```{r smear2, echo=FALSE, out.width="400px", fig.cap="MA-plot using type grouping after filtering by low expression."}
plotSmear(paired_dge.filt, lowess = TRUE, las = 1, cex.lab = 1.5, cex.axis = 1.2)
abline(h = 0, col = "blue", lwd = 2)
```

In Figure \@ref(fig:smear2) we don't see the artifact caused by the discreteness of counts at low values, where ratios between low numbers may easy lead to large fold-changes. In general, fold-changes from large expression values are more reliable than those coming from low-expression values. We don't see it because we have filtered by low-expression previously. Red line, which indicates us the tendency of the dots, shows that there is no bias, since it falls in the blue line. At the end it seems to decrease a little, but as we said, the fold-changes with a high average logCPM is more reliable. The problem would be if the biased was at the beginning. 

```{r maPlotsTumor, fig.height=36, fig.width=6, dpi=100, echo=FALSE, fig.cap="MA-plots of the tumor samples."}
par(mfrow=c(22, 3), mar=c(4, 5, 3, 1))
setmp <- paired_se.filt[, paired_se.filt$type == "tumor"]
dgetmp <- paired_dge.filt[, paired_se.filt$type == "tumor"]
for (i in 1:ncol(setmp)) {
  A <- rowMeans(assays(setmp)$logCPM)
  M <- assays(setmp)$logCPM[, i] - A
  samplename <- substr(as.character(setmp$bcr_patient_barcode[i]), 1, 12)
  smoothScatter(A, M, main=samplename, las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

We do not observe samples with major expression-level dependent biases. Although some individual samples may have little bias, in general this is corrected. Let's look now to the normal samples in Figure \@ref(fig:maPlotsTumor).

```{r maPlotsNormal, fig.height=18, fig.width=6, dpi=100, echo=FALSE, fig.cap="MA-plots of the normal samples."}
par(mfrow=c(9, 3), mar=c(4, 5, 3, 1))
setmp <- paired_se.filt[, paired_se.filt$type == "normal"]
dgetmp <- paired_dge.filt[, paired_se$type == "normal"]
for (i in 1:ncol(setmp)) {
  A <- rowMeans(assays(setmp)$logCPM)
  M <- assays(setmp)$logCPM[, i] - A
  samplename <- substr(as.character(setmp$bcr_patient_barcode[i]), 1, 12)
  smoothScatter(A, M, main=samplename, las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

We do not observe neither important expression-level dependent biases among the normal samples in Figure \@ref(fig:maPlotsNormal). There is even less bias than in tumor samples.

## Batch identification

We will search now for potential surrogate of batch effect indicators. Given that each sample names corresponds to a TCGA barcode (see https://wiki.nci.nih.gov/display/TCGA/TCGA+barcode), following the strategy described in http://bioinformatics.mdanderson.org/main/TCGABatchEffects:Overview we are going to derive different elements of the TCGA barcode and examine their distribution across samples.

```{r}
tss <- substr(colnames(paired_se.filt), 6, 7)
table(tss)
center <- substr(colnames(paired_se.filt), 27, 28)
table(center)
plate <- substr(colnames(paired_se.filt), 22, 25)
table(plate)
portionanalyte <- substr(colnames(paired_se.filt), 18, 20)
table(portionanalyte)
samplevial <- substr(colnames(paired_se.filt), 14, 16)
table(samplevial)
```

From this information we can make the following observations:

  * All samples were sequenced at the same center

  * Samples belong to three vials.

  * Samples were collected across different tissue source sites (TSS).

  * Samples were sequenced within differnt plates. There are 26 different plates.

  * Samples were sequenced using different portions and analyte combinations. There are 10 different conditions described. 
  

We are going to use the TSS as surrogate of batch effect indicator. Considering our outcome
of interest as molecular changes between sample types, tumor vs. normal, we will examine now
the cross-classification of this outcome with TSS.

```{r}
table(data.frame(TYPE=paired_se.filt$type, TSS=tss))
```

We observe that we have the same TSS in normal and tumor samples, so the TSS is not a source of variability, since they are all the same.

We can also examine the other three parameters that can lead to variability due to technical issues. We are not interested in this variability and can be a source of confussion.

```{r}
table(data.frame(TYPE=paired_se.filt$type, PORTIONALYTE=portionanalyte))
table(data.frame(TYPE=paired_se.filt$type, PLATE=plate))
table(data.frame(TYPE=paired_se.filt$type, SAMPLEVIAL=samplevial))
```

* All normal samples belong to the same vial and all tumor samples belong to the same vial. As it is remarked in the [documentation](https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/sample-type-codes), this is the correct sample vial for tumoral tissue and for normal tissue. So it does not provoque any variability. Moreover, they are all from the same aliquot (A).

* Normal samples only have two different kind of portion and analyte combinations.

* The plate is almos equal between normal and tumor.

For these reasons, the variables that may lead into batch effects they do not. 

To ilustrate how samples are grouped together by hierarchical clustering and multidimensional
scaling, we draw the Dendrogram plot and the MDS plot. We calculate again log CPM values with a higher prior count to moderate extreme fold-changes produced by low counts.

```{r sampleClustering2, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Hierarchical clustering of the samples."}
logCPM <- cpm(paired_dge.filt, log=TRUE, prior.count=3)
d <- as.dist(1-cor(logCPM, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(paired_se.filt$type))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(paired_se.filt)
names(outcome) <- colnames(paired_se.filt)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("Batch", sort(unique(batch)), levels(factor(paired_se.filt$type))), fill=sort(unique(batch)))
```

We can observe that samples cluster primarily by sample type: tumor or normal. 

In \@ref(fig:mdsPlot2) we show the corresponding MDS plot. Here we see more clearly that tumor and normal samples are separated. We can also observe that one tumor samples, corresponding to individual 8623 is in the normal cluster and also happens in the MDS plot. A closer examination of their corresponding MA-plots also reveals a slight dependence of expression changes on average expression. We may consider discarding this sample and doing the MDS plot again to have a closer look to the differences among the rest of the samples.


```{r mdsPlot2, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Multidimensional scaling plot of the samples."}
plotMDS(paired_dge.filt, labels=outcome, col=batch)
legend("bottomleft", paste("Batch", sort(unique(batch)), levels(factor(paired_se.filt$type))),
       fill=sort(unique(batch)), inset=0.05)
```

For the former case, assume that from the previous MDS plot we could decide discard that mentioned sample. But we will not do it because we have problems with the matrix size after removing this sample.

As we have not eliminate batch effects, because we have decided that there is no significant since the samples are paired and most of the technical variables are compensated or the same, we will keep working with the SummarizedExperiment object from filtered genes.

## Session information

```{r, message=FALSE}
sessionInfo()
```


